---
title: "p31662 - Boston Cohort - CytoTRACE vs CNV Correlation Analysis"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="ezRun", lib.loc=.libPaths())
    css: !expr system.file("templates/fgcz.css", package="ezRun", lib.loc=.libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Figure 5 (per-sample CNV vs stemness correlation on the Boston Xenium cohort)
```

#  {.tabset}

## Overview

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

This analysis correlates CytoTRACE2 stemness scores with Copy Number Variation (CNV) patterns
in the Boston Sullivan Xenium cohort from the p36005 Uveal Melanoma project.

**Key Questions:**
1. Does high CytoTRACE stemness correlate with specific CNV subclones?
2. Which chromosomal indels (gains/losses) are associated with stemness?
3. Is there spatial colocalization of high-stemness cells with specific CNV patterns?

**Data Sources:**
- CytoTRACE2 scores: p31662 (UM Atlas) - 14_Xenium_Stemness_Validation.Rmd
- CNV scores: 06_InSituCNV_Analysis.Rmd

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 14, fig.height = 10)

# Load required libraries
library(Seurat)
library(qs2)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(pals)
library(ggpubr)
library(viridis)

# Set paths
p31662_dir <- ".."
p36005_dir <- "../Xenium_Boston_Sullivan"

# Output directory
output_dir <- file.path(p31662_dir, "Outputs/CytoTRACE_CNV_Correlation/")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

cat("=== Analysis Setup ===\n")
cat("p31662 Atlas directory:", p31662_dir, "\n")
cat("p36005 Boston directory:", p36005_dir, "\n")
cat("Output directory:", output_dir, "\n\n")
```

## Load Data {.tabset}

### Load CytoTRACE data
```{r load-cytotrace}
cat("=== Loading CytoTRACE2 Data ===\n")

# Load Seurat object with Stemness_UCell scores (always positive, 0-1 range)
seurat_file <- file.path(p31662_dir, "Outputs/Xenium_Stemness/xenium_Boston_Sullivan_with_scores.qs2")
xenium_obj <- qs_read(seurat_file, nthreads = 8)

cat("Loaded Seurat object:", ncol(xenium_obj), "cells\n")
cat("  Samples:", paste(unique(xenium_obj$orig.ident), collapse = ", "), "\n")

# Extract metadata with UCell stemness scores
cytotrace_df <- xenium_obj@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  select(cell_id, Sample = orig.ident, CellType_Clean, Stemness_UCell) %>%
  mutate(barcode_suffix = sapply(strsplit(cell_id, "_"), function(x) x[length(x)]))

# Recompute stemness_category based on Stemness_UCell tertiles
stemness_tertiles <- quantile(cytotrace_df$Stemness_UCell, probs = c(1/3, 2/3), na.rm = TRUE)
cytotrace_df <- cytotrace_df %>%
  mutate(stemness_category = case_when(
    Stemness_UCell <= stemness_tertiles[1] ~ "Low",
    Stemness_UCell <= stemness_tertiles[2] ~ "Medium",
    TRUE ~ "High"
  ))

cat("Extracted metadata:", nrow(cytotrace_df), "cells\n")
cat("  Stemness_UCell range:", round(min(cytotrace_df$Stemness_UCell, na.rm = TRUE), 3), "-",
    round(max(cytotrace_df$Stemness_UCell, na.rm = TRUE), 3), "\n")
cat("  Stemness tertiles (Low/Medium/High cutoffs):",
    round(stemness_tertiles[1], 4), ",", round(stemness_tertiles[2], 4), "\n")
cat("  Stemness categories:", paste(names(table(cytotrace_df$stemness_category)),
    table(cytotrace_df$stemness_category), collapse = ", "), "\n")
```

### Load CNV data from h5ad files (recalculate from X_cnv matrix)
```{r load-cnv}
cat("\n=== Loading CNV Data from h5ad files ===\n")

# The CSV cnv_score is incorrect (constant values). Need to recalculate from X_cnv matrix.
library(anndataR)

cnv_output_dir <- file.path(p36005_dir, "06_InSituCNV_Outputs")
cnv_h5ad_files <- list.files(cnv_output_dir, pattern = "_cnv\\.h5ad$", full.names = TRUE)

cat("Found", length(cnv_h5ad_files), "CNV h5ad files\n")

# Function to calculate CNV score from X_cnv matrix
# Using RMS (root mean square) of CNV values as overall CNV burden
calculate_cnv_score <- function(X_cnv) {
  # Convert to matrix if needed
  if (inherits(X_cnv, "dgRMatrix") || inherits(X_cnv, "dgCMatrix")) {
    X_mat <- as.matrix(X_cnv)
  } else {
    X_mat <- as(X_cnv, "matrix")
  }
  # Calculate RMS of each row (cell) - represents overall CNV deviation
  sqrt(rowMeans(X_mat^2))
}

cnv_list <- list()
for (h5ad_file in cnv_h5ad_files) {
  sample_name <- basename(h5ad_file) %>% gsub("_cnv\\.h5ad$", "", .)
  cat("  Processing:", sample_name, "\n")

  # Load h5ad file
  adata <- read_h5ad(h5ad_file)

  # Get cell barcodes
  cell_barcodes <- rownames(adata$obs)

  # Recalculate CNV score from X_cnv matrix
  X_cnv <- adata$obsm$X_cnv
  cnv_score_recalc <- calculate_cnv_score(X_cnv)

  # Get other metadata
  cnv_data <- data.frame(
    cell_type = adata$obs$cell_type,
    cnv_score = cnv_score_recalc,
    x_centroid = adata$obs$x_centroid,
    y_centroid = adata$obs$y_centroid,
    sample = sample_name,
    row.names = cell_barcodes
  )

  cnv_list[[sample_name]] <- cnv_data

  cat("    Cells:", nrow(cnv_data), "\n")
  cat("    CNV score range:", round(min(cnv_score_recalc), 4), "-",
      round(max(cnv_score_recalc), 4), "\n")
}

cnv_all <- bind_rows(cnv_list, .id = "sample_source")
cat("\nTotal CNV records:", nrow(cnv_all), "\n")
cat("Recalculated CNV score summary:\n")
print(summary(cnv_all$cnv_score))
```

### Merge datasets
```{r merge-datasets}
cat("\n=== Merging Datasets ===\n")

# The CNV data has cell barcodes as rownames (e.g., "aaaaihba-1")
# The cytotrace_df has full cell IDs (e.g., "S22-19XXX-A1_Slide12_aaaaihba-1")
# Strategy: barcode_suffix already extracted in previous chunk

cnv_all$barcode_suffix <- rownames(cnv_all)

# Merge by barcode suffix
merged_df <- left_join(cytotrace_df, cnv_all,
                       by = "barcode_suffix",
                       suffix = c("_cyto", "_cnv"))

cat("Merge results:\n")
cat("  Stemness (UCell) cells:", nrow(cytotrace_df), "\n")
cat("  CNV records:", nrow(cnv_all), "\n")
cat("  Merged cells with CNV data:", sum(!is.na(merged_df$cnv_score)), "\n")

# Create analysis dataframe with complete cases
analysis_df <- merged_df %>%
  filter(!is.na(Stemness_UCell), !is.na(cnv_score))

cat("  Cells with BOTH scores:", nrow(analysis_df), "\n")

# Order stemness_category factor correctly
analysis_df$stemness_category <- factor(analysis_df$stemness_category,
                                         levels = c("Low", "Medium", "High"))

# Display score ranges
cat("\nScore ranges:\n")
cat("  Stemness_UCell (CytoTRACE2 signature):", round(min(analysis_df$Stemness_UCell), 3), "-",
    round(max(analysis_df$Stemness_UCell), 3), "\n")
cat("  CNV score (recalculated RMS):", round(min(analysis_df$cnv_score), 4), "-",
    round(max(analysis_df$cnv_score), 4), "\n")

# Check sample distribution
cat("\nSamples in analysis:\n")
print(table(analysis_df$Sample))
```

## Data Summary {.tabset}

### Summary statistics
```{r summary-stats}
cat("=== Data Summary ===\n\n")

# By sample
sample_summary <- analysis_df %>%
  group_by(sample) %>%
  summarise(
    n_cells = n(),
    mean_stemness = mean(Stemness_UCell),
    median_stemness = median(Stemness_UCell),
    mean_cnv = mean(cnv_score),
    median_cnv = median(cnv_score),
    .groups = "drop"
  )

print(sample_summary)

# By stemness category
stem_summary <- analysis_df %>%
  group_by(stemness_category) %>%
  summarise(
    n_cells = n(),
    mean_cnv = mean(cnv_score),
    median_cnv = median(cnv_score),
    sd_cnv = sd(cnv_score),
    .groups = "drop"
  ) %>%
  arrange(stemness_category)

cat("\nBy stemness category:\n")
print(stem_summary)
```

## Correlation Analysis {.tabset}

### Scatter plot: Stemness vs CNV score
```{r scatter-stemness-cnv, fig.width=14, fig.height=10}
# Overall correlation
p_overall <- ggplot(analysis_df, aes(x = Stemness_UCell, y = cnv_score)) +
  geom_point(alpha = 0.2, size = 0.8, color = "darkgray") +
  geom_smooth(method = "lm", color = "red", se = TRUE, linewidth = 1) +
  stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.95,
           aes(label = ..r.label..), size = 5) +
  stat_cor(method = "spearman", label.x.npc = 0.05, label.y.npc = 0.85,
           aes(label = ..r.label..), size = 4) +
  theme_minimal() +
  labs(title = "All Cells: CytoTRACE2 Stemness (UCell) vs CNV Score",
       x = "Stemness Score (CytoTRACE2 UCell)",
       y = "CNV Score (RMS)")

print(p_overall)
ggsave(file.path(output_dir, "scatter_stemness_vs_cnv_overall.png"),
       p_overall, width = 10, height = 8, dpi = 300, bg = "white")

# By sample
p_by_sample <- ggplot(analysis_df, aes(x = Stemness_UCell, y = cnv_score, color = sample)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  facet_wrap(~sample, scales = "free", ncol = 3) +
  stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.95, size = 3.5) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(title = "All Cells: CytoTRACE2 Stemness (UCell) vs CNV Score by Sample",
       x = "Stemness Score (CytoTRACE2 UCell)",
       y = "CNV Score (RMS)",
       color = "Sample")

print(p_by_sample)
ggsave(file.path(output_dir, "scatter_stemness_vs_cnv_by_sample.png"),
       p_by_sample, width = 14, height = 10, dpi = 300, bg = "white")

# Calculate correlation statistics
cat("\n=== Correlation Statistics ===\n")
cor_pearson <- cor(analysis_df$Stemness_UCell, analysis_df$cnv_score, method = "pearson")
cor_spearman <- cor(analysis_df$Stemness_UCell, analysis_df$cnv_score, method = "spearman")

cat("Pearson correlation:", round(cor_pearson, 4), "\n")
cat("Spearman correlation:", round(cor_spearman, 4), "\n")

cor_test <- cor.test(analysis_df$Stemness_UCell, analysis_df$cnv_score, method = "pearson")
cat("p-value:", format(cor_test$p.value, scientific = TRUE), "\n")

# By sample correlations
cat("\nBy sample:\n")
by_sample_cor <- analysis_df %>%
  group_by(sample) %>%
  summarise(
    n = n(),
    pearson_r = cor(Stemness_UCell, cnv_score, use = "complete.obs"),
    p_value = cor.test(Stemness_UCell, cnv_score)$p.value,
    .groups = "drop"
  )
print(by_sample_cor)
```

### Stemness categories vs CNV categories
```{r stemness-categories-cnv, fig.width=14, fig.height=10}
# Create CNV categories (high/low based on median)
cnv_median <- median(analysis_df$cnv_score)
analysis_df$cnv_category <- ifelse(analysis_df$cnv_score > cnv_median, "High_CNV", "Low_CNV")

# Contingency table
contingency <- table(stemness = analysis_df$stemness_category,
                     cnv = analysis_df$cnv_category)

cat("=== Stemness Category vs CNV Category ===\n")
print(contingency)

# Chi-square test
chi_test <- chisq.test(contingency)
cat("\nChi-square test:\n")
cat("  X-squared:", round(chi_test$statistic, 4), "\n")
cat("  p-value:", format(chi_test$p.value, scientific = TRUE), "\n")

# Bar plot
contingency_df <- as.data.frame(contingency)
colnames(contingency_df) <- c("Stemness", "CNV", "Count")

p_bar <- ggplot(contingency_df, aes(x = Stemness, y = Count, fill = CNV)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9),
            vjust = -0.25, size = 4) +
  scale_fill_manual(values = c("Low_CNV" = "#2c7bb6", "High_CNV" = "#d7191c")) +
  theme_minimal() +
  labs(title = "Stemness Category vs CNV Category",
       x = "Stemness Category (CytoTRACE2)",
       y = "Number of Cells",
       fill = "CNV Category")

print(p_bar)
ggsave(file.path(output_dir, "bar_stemness_vs_cnv_category.png"),
       p_bar, width = 10, height = 8, dpi = 300, bg = "white")

# Box plot: CNV score by stemness category
p_box <- ggplot(analysis_df, aes(x = stemness_category, y = cnv_score, fill = stemness_category)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
  geom_violin(alpha = 0.3) +
  scale_fill_manual(values = c("Low" = "#2166AC", "Medium" = "#999999", "High" = "#B2182B")) +
  stat_compare_means(method = "kruskal.test", label.y.npc = 0.95) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "CNV Score Distribution by Stemness Category",
       x = "Stemness Category (CytoTRACE2)",
       y = "CNV Score")

print(p_box)
ggsave(file.path(output_dir, "boxplot_cnv_by_stemness_category.png"),
       p_box, width = 10, height = 8, dpi = 300, bg = "white")
```

### Kruskal-Wallis test and pairwise comparisons
```{r pairwise-tests}
cat("\n=== Pairwise Comparisons ===\n")

# Kruskal-Wallis test
kw_test <- kruskal.test(cnv_score ~ stemness_category, data = analysis_df)
cat("Kruskal-Wallis test:\n")
cat("  chi-squared:", round(kw_test$statistic, 4), "\n")
cat("  p-value:", format(kw_test$p.value, scientific = TRUE), "\n")

# Pairwise Wilcoxon tests
pairwise_test <- pairwise.wilcox.test(analysis_df$cnv_score,
                                      analysis_df$stemness_category,
                                      p.adjust.method = "BH")
cat("\nPairwise Wilcoxon tests (BH adjusted):\n")
print(pairwise_test$p.value)
```

## Cell Type Analysis {.tabset}

### Stemness and CNV by cell type
```{r celltype-analysis}
cat("\n=== Analysis by Cell Type ===\n\n")

# Summary by cell type
celltype_summary <- analysis_df %>%
  group_by(CellType_Clean) %>%
  summarise(
    n_cells = n(),
    mean_stemness = mean(Stemness_UCell),
    median_stemness = median(Stemness_UCell),
    mean_cnv = mean(cnv_score),
    median_cnv = median(cnv_score),
    pct_high_stemness = sum(stemness_category == "High") / n() * 100,
    pct_high_cnv = sum(cnv_category == "High_CNV") / n() * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(mean_stemness))

print(celltype_summary)

# Calculate standard error for each cell type
celltype_se <- analysis_df %>%
  group_by(CellType_Clean) %>%
  summarise(
    se_stemness = sd(Stemness_UCell) / sqrt(n()),
    .groups = "drop"
  )

# Add SE to summary
celltype_summary <- celltype_summary %>%
  left_join(celltype_se, by = "CellType_Clean")

# Plot: Stemness by cell type
p_cell_stem <- ggplot(celltype_summary, aes(x = reorder(CellType_Clean, mean_stemness), y = mean_stemness)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_errorbar(aes(ymin = mean_stemness - se_stemness, ymax = mean_stemness + se_stemness),
                width = 0.2) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Mean Stemness Score by Cell Type",
       x = "Cell Type",
       y = "Mean CytoTRACE2 Stemness Score")

print(p_cell_stem)
ggsave(file.path(output_dir, "barplot_stemness_by_celltype.png"),
       p_cell_stem, width = 12, height = 8, dpi = 300, bg = "white")

# Plot: CNV by cell type
p_cell_cnv <- ggplot(celltype_summary, aes(x = reorder(CellType_Clean, mean_cnv), y = mean_cnv)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Mean CNV Score by Cell Type",
       x = "Cell Type",
       y = "Mean CNV Score")

print(p_cell_cnv)
ggsave(file.path(output_dir, "barplot_cnv_by_celltype.png"),
       p_cell_cnv, width = 12, height = 8, dpi = 300, bg = "white")
```

### Melanoma cells: High stemness vs High CNV
```{r melanoma-analysis}
# Focus on melanoma cells
melanoma_df <- analysis_df %>%
  filter(grepl("melanoma|Melanoma|tumor|Tumor", CellType_Clean, ignore.case = TRUE))

if (nrow(melanoma_df) > 50) {
  cat("\n=== Melanoma Cell Analysis ===\n")
  cat("Melanoma cells:", nrow(melanoma_df), "\n")

  # Overlap analysis
  high_stem <- melanoma_df$stemness_category == "High"
  high_cnv <- melanoma_df$cnv_category == "High_CNV"

  both_high <- sum(high_stem & high_cnv)
  stem_high_cnv_low <- sum(high_stem & !high_cnv)
  stem_low_cnv_high <- sum(!high_stem & high_cnv)

  cat("\nMelanoma cell overlaps:\n")
  cat("  High stemness + High CNV:", both_high, "\n")
  cat("  High stemness + Low CNV:", stem_high_cnv_low, "\n")
  cat("  Low stemness + High CNV:", stem_low_cnv_high, "\n")

  # Expected vs observed for high stemness + high CNV
  if (both_high > 0) {
    pct_high_stem <- sum(high_stem) / nrow(melanoma_df) * 100
    pct_high_cnv <- sum(high_cnv) / nrow(melanoma_df) * 100
    pct_both <- both_high / nrow(melanoma_df) * 100
    expected_pct <- pct_high_stem * pct_high_cnv / 100

    cat("\nExpected overlap (if independent):", round(expected_pct, 1), "%\n")
    cat("Observed overlap:", round(pct_both, 1), "%\n")
    cat("Fold enrichment:", round(pct_both / expected_pct, 2), "\n")
  }
}
```

### Scatter plots: Melanoma cells only
```{r melanoma-scatter-plots, fig.width=14, fig.height=10}
# Check if enough melanoma cells exist
if (nrow(melanoma_df) > 50) {
  cat("\n=== Melanoma Cell Scatter Plots ===\n")
  cat("Total melanoma cells:", nrow(melanoma_df), "\n")

  # 1. Overall melanoma correlation plot
  p_melanoma_overall <- ggplot(melanoma_df, aes(x = Stemness_UCell, y = cnv_score)) +
    geom_point(alpha = 0.2, size = 0.8, color = "darkgray") +
    geom_smooth(method = "lm", color = "red", se = TRUE, linewidth = 1) +
    stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.95,
             aes(label = ..r.label..), size = 5) +
    stat_cor(method = "spearman", label.x.npc = 0.05, label.y.npc = 0.85,
             aes(label = ..r.label..), size = 4) +
    theme_minimal() +
    labs(title = "MELANOMA CELLS ONLY: CytoTRACE2 Stemness (UCell) vs CNV Score",
         x = "Stemness Score (CytoTRACE2 UCell)",
         y = "CNV Score (RMS)")

  print(p_melanoma_overall)
  ggsave(file.path(output_dir, "scatter_stemness_vs_cnv_melanoma_overall.png"),
         p_melanoma_overall, width = 10, height = 8, dpi = 300, bg = "white")

  # 2. By-sample melanoma correlation plots
  p_melanoma_by_sample <- ggplot(melanoma_df, aes(x = Stemness_UCell, y = cnv_score, color = sample)) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
    facet_wrap(~sample, scales = "free", ncol = 3) +
    stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.95, size = 3.5) +
    theme_minimal() +
    theme(legend.position = "bottom") +
    labs(title = "MELANOMA CELLS ONLY: CytoTRACE2 Stemness (UCell) vs CNV Score by Sample",
         x = "Stemness Score (CytoTRACE2 UCell)",
         y = "CNV Score (RMS)",
         color = "Sample")

  print(p_melanoma_by_sample)
  ggsave(file.path(output_dir, "scatter_stemness_vs_cnv_melanoma_by_sample.png"),
         p_melanoma_by_sample, width = 14, height = 10, dpi = 300, bg = "white")

  # 3. Correlation statistics for melanoma cells
  cat("\n=== Melanoma Cell Correlation Statistics ===\n")
  cor_pearson_mel <- cor(melanoma_df$Stemness_UCell, melanoma_df$cnv_score, method = "pearson")
  cor_spearman_mel <- cor(melanoma_df$Stemness_UCell, melanoma_df$cnv_score, method = "spearman")

  cat("Overall (melanoma cells only):\n")
  cat("  Pearson correlation:", round(cor_pearson_mel, 4), "\n")
  cat("  Spearman correlation:", round(cor_spearman_mel, 4), "\n")

  cor_test_mel <- cor.test(melanoma_df$Stemness_UCell, melanoma_df$cnv_score, method = "pearson")
  cat("  p-value:", format(cor_test_mel$p.value, scientific = TRUE), "\n")

  # By sample correlations for melanoma cells
  cat("\nBy sample (melanoma cells only):\n")
  by_sample_cor_mel <- melanoma_df %>%
    group_by(sample) %>%
    summarise(
      n_melanoma = n(),
      pearson_r = cor(Stemness_UCell, cnv_score, use = "complete.obs"),
      p_value = cor.test(Stemness_UCell, cnv_score)$p.value,
      .groups = "drop"
    )
  print(by_sample_cor_mel)

} else {
  cat("Not enough melanoma cells for scatter plots (n =", nrow(melanoma_df), ")\n")
}
```

## Spatial Analysis {.tabset}

### Spatial plots: Stemness and CNV
```{r spatial-plots, fig.width=18, fig.height=6}
cat("\n=== Generating Spatial Plots ===\n")

# Check if we have coordinates
if (all(c("x_centroid", "y_centroid") %in% colnames(analysis_df))) {
  cat("Spatial coordinates available. Generating plots...\n")

  # Get unique samples
  samples <- unique(analysis_df$sample)

  for (sample_name in samples) {
    cat("\nSample:", sample_name, "\n")

    sample_df <- analysis_df %>%
      filter(sample == sample_name)

    # Stemness score spatial plot
    p1 <- ggplot(sample_df, aes(x = x_centroid, y = y_centroid, color = Stemness_UCell)) +
      geom_point(size = 0.5, alpha = 0.7) +
      scale_color_viridis_c(option = "magma", name = "Stemness") +
      coord_fixed() +
      theme_minimal() +
      theme(legend.position = "right") +
      labs(title = paste("CytoTRACE2 Stemness -", sample_name),
           x = "X (microns)", y = "Y (microns)")

    # CNV score spatial plot
    p2 <- ggplot(sample_df, aes(x = x_centroid, y = y_centroid, color = cnv_score)) +
      geom_point(size = 0.5, alpha = 0.7) +
      scale_color_gradient2(low = "blue", mid = "white", high = "red",
                           midpoint = median(sample_df$cnv_score, na.rm = TRUE),
                           name = "CNV") +
      coord_fixed() +
      theme_minimal() +
      theme(legend.position = "right") +
      labs(title = paste("CNV Score -", sample_name),
           x = "X (microns)", y = "Y (microns)")

    # Stemness category spatial plot
    stem_colors <- c("Low" = "#2166AC", "Medium" = "#999999", "High" = "#B2182B")
    p3 <- ggplot(sample_df, aes(x = x_centroid, y = y_centroid, color = stemness_category)) +
      geom_point(size = 0.5, alpha = 0.7) +
      scale_color_manual(values = stem_colors, name = "Stemness") +
      coord_fixed() +
      theme_minimal() +
      theme(legend.position = "right") +
      labs(title = paste("Stemness Category -", sample_name),
           x = "X (microns)", y = "Y (microns)")

    # Combined plot
    p_combined <- p1 | p2 | p3
    print(p_combined)

    # Save
    safe_name <- gsub("[^a-zA-Z0-9_-]", "_", sample_name)
    ggsave(file.path(output_dir, paste0("spatial_stemness_cnv_", safe_name, ".png")),
           p_combined, width = 18, height = 6, dpi = 300, bg = "white")
  }

} else {
  cat("Spatial coordinates not available in the merged dataset.\n")
}
```

### Spatial plot: Combined Stemness x CNV categories
```{r spatial-combined, fig.width=12, fig.height=10}
if (all(c("x_centroid", "y_centroid") %in% colnames(analysis_df))) {

  # Create combined category
  analysis_df$stemness_cnv_combined <- paste(analysis_df$stemness_category,
                                              analysis_df$cnv_category,
                                              sep = "_\n")

  # Define colors for combined categories
  combined_colors <- c(
    "Low\nLow_CNV" = "#deebf7",
    "Low\nHigh_CNV" = "#6baed6",
    "Medium\nLow_CNV" = "#e5e5e5",
    "Medium\nHigh_CNV" = "#fd8d3c",
    "High\nLow_CNV" = "#fee5d9",
    "High\nHigh_CNV" = "#d7191c"
  )

  # Plot for each sample
  for (sample_name in unique(analysis_df$sample)) {
    sample_df <- analysis_df %>%
      filter(sample == sample_name)

    p_combined <- ggplot(sample_df, aes(x = x_centroid, y = y_centroid, color = stemness_cnv_combined)) +
      geom_point(size = 0.8, alpha = 0.8) +
      scale_color_manual(values = combined_colors, name = "Stemness x CNV") +
      coord_fixed() +
      theme_minimal() +
      theme(legend.position = "right") +
      labs(title = paste("Stemness x CNV Combined -", sample_name),
           x = "X (microns)", y = "Y (microns)")

    print(p_combined)

    safe_name <- gsub("[^a-zA-Z0-9_-]", "_", sample_name)
    ggsave(file.path(output_dir, paste0("spatial_combined_categories_", safe_name, ".png")),
           p_combined, width = 12, height = 10, dpi = 300, bg = "white")
  }
}
```

## Classical UM Chromosome Aberrations {.tabset}

### High stemness vs high CNV analysis
```{r um-aberrations}
cat("\n=== Classical UM Chromosome Aberrations Context ===\n")
cat("Note: Chromosome-level analysis requires gene-level CNV data from h5ad files.\n")
cat("Current analysis uses aggregate CNV scores from infercnvpy.\n")

cat("\nFor this analysis, CNV scores reflect overall genomic instability.\n")
cat("Higher CNV scores indicate more chromosomal aberrations (gains/losses).\n")

# Define quartiles
stem_q75 <- quantile(analysis_df$Stemness_UCell, 0.75)
stem_q25 <- quantile(analysis_df$Stemness_UCell, 0.25)
cnv_q75 <- quantile(analysis_df$cnv_score, 0.75)
cnv_q25 <- quantile(analysis_df$cnv_score, 0.25)

# Create quartile groups
analysis_df$stem_quartile <- cut(analysis_df$Stemness_UCell,
                                 breaks = c(-Inf, stem_q25, stem_q75, Inf),
                                 labels = c("Q1 (Low)", "Q2-Q3 (Mid)", "Q4 (High)"))
analysis_df$cnv_quartile <- cut(analysis_df$cnv_score,
                                breaks = c(-Inf, cnv_q25, cnv_q75, Inf),
                                labels = c("Q1 (Low)", "Q2-Q3 (Mid)", "Q4 (High)"))

# Crosstable
quartile_table <- table(stem = analysis_df$stem_quartile,
                        cnv = analysis_df$cnv_quartile)
cat("\nQuartile crosstab:\n")
print(quartile_table)

# Chi-square test on quartiles
quartile_test <- chisq.test(quartile_table)
cat("\nChi-square test (quartiles):\n")
cat("  p-value:", format(quartile_test$p.value, scientific = TRUE), "\n")
```

### Key findings summary
```{r key-findings}
cat("\n=== Key Findings Summary ===\n")

# Proportion of high stemness cells with high CNV
n_high_stem <- sum(analysis_df$stemness_category == "High")
n_high_stem_high_cnv <- sum(analysis_df$stemness_category == "High" &
                             analysis_df$cnv_category == "High_CNV")
pct_high_stem_high_cnv <- 100 * n_high_stem_high_cnv / n_high_stem

cat("\n1. High Stemness and CNV:\n")
cat("   High stemness cells:", n_high_stem, "\n")
cat("   High stemness + High CNV:", n_high_stem_high_cnv,
    sprintf("(%.1f%%)\n", pct_high_stem_high_cnv))

# Proportion of low stemness cells with high CNV
n_low_stem <- sum(analysis_df$stemness_category == "Low")
n_low_stem_high_cnv <- sum(analysis_df$stemness_category == "Low" &
                           analysis_df$cnv_category == "High_CNV")
pct_low_stem_high_cnv <- 100 * n_low_stem_high_cnv / n_low_stem

cat("\n2. Low Stemness and CNV:\n")
cat("   Low stemness cells:", n_low_stem, "\n")
cat("   Low stemness + High CNV:", n_low_stem_high_cnv,
    sprintf("(%.1f%%)\n", pct_low_stem_high_cnv))

# Correlation interpretation
cat("\n3. Correlation Interpretation:\n")
if (cor_test$p.value < 0.05) {
  if (cor_pearson > 0) {
    cat("   Significant POSITIVE correlation (p <", format(cor_test$p.value, digits = 2), ")\n")
    cat("   Higher stemness tends to associate with higher CNV burden\n")
  } else {
    cat("   Significant NEGATIVE correlation (p <", format(cor_test$p.value, digits = 2), ")\n")
    cat("   Higher stemness tends to associate with lower CNV burden\n")
  }
} else {
  cat("   No significant correlation (p =", format(cor_test$p.value, digits = 3), ")\n")
  cat("   Stemness and CNV appear independent in this dataset\n")
}
```

## Export Results {.tabset}

### Export combined metadata
```{r export-results}
cat("\n=== Exporting Results ===\n")

# Export the merged dataframe
export_df <- analysis_df %>%
  select(cell_id, barcode_suffix, Sample, CellType_Clean,
         Stemness_UCell, stemness_category, cnv_score, cnv_category,
         sample, x_centroid, y_centroid)

output_file <- file.path(output_dir, "cytotrace_cnv_boston_combined.csv")
write.csv(export_df, output_file, row.names = FALSE)
cat("Exported to:", output_file, "\n")
cat("  Rows:", nrow(export_df), "\n")
cat("  Columns:", ncol(export_df), "\n")
```

### Summary report
```{r summary-report}
cat("\n=== Analysis Summary ===\n")
cat("Completed at:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n\n")

cat("Dataset: Boston Sullivan cohort (Xenium)\n")
cat("  Total cells analyzed:", nrow(analysis_df), "\n")
cat("  Samples:", length(unique(analysis_df$sample)), "\n")
cat("  Cell types:", length(unique(analysis_df$CellType_Clean)), "\n\n")

cat("Key Results:\n")
cat("  1. Pearson correlation:", round(cor_pearson, 4), "\n")
cat("  2. Spearman correlation:", round(cor_spearman, 4), "\n")
cat("  3. p-value:", format(cor_test$p.value, scientific = TRUE), "\n")
cat("  4. High stemness + High CNV:", sprintf("%.1f%%", pct_high_stem_high_cnv), "\n")
cat("  5. Low stemness + High CNV:", sprintf("%.1f%%", pct_low_stem_high_cnv), "\n")

cat("\nOutput files:\n")
cat("  - CSV: cytotrace_cnv_boston_combined.csv\n")
cat("  - Plots: Various PNG files in", output_dir, "\n")
```

## Session Info

```{r session-info}
sessionInfo()
```
